---
title: "Data analysis of Intelligence Squared events"
output: html_document
---


# Data analysis on gender distribution among speakers and chairs in Intelligence Squared events from the beginning to 23/03/2015

## Initialization

Load libraries

```{r load libraries}
library(gdata)
library(lattice)
library(vcd)
library(reshape2)
library(dplyr)
```

Read input file:

```{r}
df = read.xls("U:\\my research\\Gender equality\\Intelligence Squared\\data\\dataset_25012016.xlsx", sheet = "edited", header = TRUE, colClasses = c("Date", "character", "character", "character", "character", "factor", "factor", "factor", "factor", "factor", "factor", "factor"))
#df[,"date"] = as.Date(df[,"date"], format="%Y-%m-%d")
# number of male/female speakers/chairs:
df[,2:5] = sapply(df[,2:5], as.numeric) 
# binary topic variables:
df[7:ncol(df)] = lapply(df[,7:ncol(df)], function(x) x <- sub("^$", 0,x)) 
df[,7:ncol(df)] = sapply(sapply(df[,7:ncol(df)], as.numeric), as.logical)
```


Convert dataset into dataframe so that each row corresponds to a person:

```{r}
df$ID = as.factor(rownames(df))
df1 = melt(df[,(colnames(df) %in% c("date", "female.speakers", "male.speakers","female.chair","male.chair","ID"))], id=c("ID", "date"))
df1$gender = as.factor(ifelse(grepl("f.",df1$variable),"female", "male"))
df1$type = as.factor(ifelse(grepl(".sp",df1$variable),"speaker", "chair"))
df1$ID = as.factor(df1$ID)
df2 = df1[df1$value>0,]
colnames(df2)[colnames(df2)=="value"] = "freq"
df2$ID = as.factor(df2$ID)
head(df2)
```

Overview of dataset:

```{r}
summary(df)
plot(df)
summary(df2)
plot(df2)
```

## Measures of gender balance

Define the (proportional) difference between female and male speakers in each event: 

```{r}
df$gender_difference = df$male.speakers-df$female.speakers
summary(df$gender_difference)
table(df$gender_difference)
table(sign(df$gender_difference))
prop.table(table(sign(df$gender_difference)))
hist(df$gender_difference, main = "Difference in number of males vs. females per event", xlab = "Difference (absolute values)", col = "grey")
df$gender_ratio = ifelse(df$female.speakers == 0, 0, round(df$male.speakers/df$female.speakers,2))
summary(df$gender_ratio)
hist(df$gender_ratio, main = "Ratio of males vs. females per event", xlab = "Ratio", col = "grey")
table(df$gender_ratio)
table(sign(df$gender_ratio))
```

### Total number of women over all people

Number of female speakers/chairs and male speakers/chairs:

```{r}
gender.type = xtabs(freq ~ gender + type, data = df2)
barplot(gender.type, legend=T, main = "Total number of speakers and chairs by gender")
boxplot(freq ~ gender, data = df2)
boxplot(freq ~ gender+type, data = df2)
```


```{r}
rowSums(gender.type)[1]/sum(gender.type)
```

Of all people (speakers + chairs), `r rowSums(gender.type)[1]` (`r 100*round(rowSums(gender.type)[1]/sum(gender.type),2)`%) are women and `r rowSums(gender.type)[2]` (`r 100*round(rowSums(gender.type)[2]/sum(gender.type),2)`%) are men.

### Gender distribution by type

```{r}
(gender.type)
prop.table(gender.type,1)
```

Of all women, `r gender.type[1,1]` (`r 100*round(prop.table(gender.type,1)[1,1],2)`%) are chairs and  `r gender.type[1,2]` (`r 100*round(prop.table(gender.type,1)[1,2],2)`%) are speakers; of all men, `r gender.type[2,1]` (`r 100*round(prop.table(gender.type,1)[2,1],2)`%) are chairs and `r gender.type[2,2]` (`r 100*round(prop.table(gender.type,1)[2,2],2)`%) are speakers. 

```{r}
(gender.type)
prop.table(gender.type,2)
```

`r gender.type[1,1]` chairs are female (`r 100*round(prop.table(gender.type,2)[1,1], 2)`%) of all chairs, `r gender.type[2,1]` (`r 100*round(prop.table(gender.type,2)[2,1], 2)`%) are male; `r 100*round(prop.table(gender.type, 2)[1,2], 2)`% of speakers are female, `r 100*round(prop.table(gender.type,2)[2,2], 2)`% are male.

```{r}
prop.table(gender.type,1)
```


I test the null hypothesis that the proportions of females in each group (chairs and speakers) are the same:

```{r}
prop.test(gender.type[,1],gender.type[,1]+gender.type[,2])
prop.table(gender.type,1)
assocplot(gender.type)
```

This shows, as expected, that the proportion of female chairs is significantly lower than the proportion of male chairs.


The same holds for speakers:


```{r}
prop.test(gender.type[,2],gender.type[,1]+gender.type[,2])
prop.table(gender.type,1)
assocplot(gender.type)
```

Is there a significant difference in the gender distribution between chairs and speakers?

```{r}
(gender.type)
chisq.test(gender.type)
assocstats(gender.type)
assocplot(gender.type)
```

This shows that there is a statistically significant difference between the gender distribution of chairs and that of speakers, and the effect of this is small. 
In other words, women are significantly under-represented among chairs and speakers, compared to men, and that the difference is larger for speakers.

```{r}
prop.test(gender.type[1,1], colSums(gender.type)[1], p=prop.table(gender.type,2)[1,2])
```

There are significantly more female chairs than would be expected given the proportion of female speakers out of all speakers.
This can be explained by the fact that an event only needs one chair, so even if the panel of speakers is all-male, it's relatively more likely that the chair is female. 

Let's check this.

## All-male speakers and female chairs

```{r}
all_male_speakers = droplevels(filter(df, female.speakers == 0 & male.speakers > 0))
all_male_speakers_chaired = droplevels(filter(all_male_speakers, male.chair==1 | female.chair==1))
all_male_speakers_chaired_female_chair = prop.table(table(all_male_speakers_chaired$female.chair))
all_events_chaired = droplevels(filter(df, female.chair == 1 | male.chair == 1))
all_events_chaired_female_chair = prop.table(table(all_events_chaired$female.chair))
all_events_chaired_male_chair = prop.table(table(all_events_chaired$male.chair))
```

There are `r nlevels(all_male_speakers$ID)` events with an all-male panel; this corresponds to `r 100*round(nlevels(all_male_speakers$ID)/nlevels(df$ID), 2)` of all events.
Of such events, `r nlevels(all_male_speakers_chaired$ID)` have a chair.

Of such chaired all-male-speaker events, `r table(all_male_speakers_chaired$female.chair)[2]` (`r 100*round(all_male_speakers_chaired_female_chair[2],2)`%) have a female chair. 

Of all chaired events (`r nlevels(all_events_chaired$ID)`), `r table(all_events_chaired$female.chair)[2]` (`r 100*round(all_events_chaired_female_chair[2], 2)`%) have a female chair. 

Is this difference statistically significant?

```{r}
prop.test(matrix(c(table(all_male_speakers_chaired$female.chair)[2],nlevels(all_male_speakers_chaired$ID)-table(all_male_speakers_chaired$female.chair)[2],table(all_events_chaired$female.chair)[2],nlevels(all_events_chaired$ID)-table(all_events_chaired$female.chair)[2]), nrow=2, byrow=T))
```

So, it's not true that when the speakers are all male, we're more likely to have a female chair.


### Proportion of women 

There are `r nrow(filter(df, female.speakers+female.chair>male.speakers+male.chair))` events (`r 100*round(nrow(filter(df, female.speakers+female.chair>male.speakers+male.chair))/nrow(df),2)`%) where the number of females (speakers + chairs) is higher than the number of males (speakers + chairs); `r nrow(filter(df, female.speakers+female.chair==male.speakers+male.chair))` events (`r 100*round(nrow(filter(df, female.speakers+female.chair==male.speakers+male.chair))/nrow(df),2)`%) where the number of females (speakers + chairs) is the same as the number of males (speakers + chairs); `r nrow(filter(df, female.speakers+female.chair<male.speakers+male.chair))` events (`r 100*round(nrow(filter(df, female.speakers+female.chair<male.speakers+male.chair))/nrow(df), 2)`%) where the number of females (speakers + chairs) is lower than the number of males (speakers + chairs).

There are `r nrow(filter(df, female.speakers>male.speakers))` events (`r 100*round(nrow(filter(df, female.speakers>male.speakers))/nrow(df),2)`%) where the number of females (speakers) is higher than the number of males (speakers); `r nrow(filter(df, female.speakers==male.speakers))` events (`r 100*round(nrow(filter(df, female.speakers==male.speakers))/nrow(df), 2)`%) where the number of females (speakers) is the same as the number of males (speakers); `r nrow(filter(df, female.speakers<male.speakers))` events (`r 100*round(nrow(filter(df, female.speakers<male.speakers))/nrow(df),2)`%) where the number of females (speakers) is lower than the number of males (speakers).

There are `r nrow(filter(df, female.chair>male.chair))` events (`r 100*round(nrow(filter(df, female.chair>male.chair))/nrow(df),2)`%) where the number of females (chairs) is higher than the number of males (chairs); `r nrow(filter(df, female.chair==male.chair))` events (`r 100*round(nrow(filter(df, female.chair==male.chair))/nrow(df),2)`%) where the number of females (chairs) is the same as the number of males (chairs); `r nrow(filter(df, female.chair<male.chair))` events (`r 100*round(nrow(filter(df, female.chair<male.chair))/nrow(df),2)`%) where the number of females (chairs) is lower than the number of males (chairs).

`r nrow(filter(df, female.speakers == 0))` (`r 100*round(nrow(filter(df, female.speakers == 0))/nrow(df),2)`%) have no female speakers and `r nrow(filter(df, female.speakers == 0 & female.chair == 0))` (`r 100*round(nrow(filter(df, female.speakers == 0 & female.chair == 0))/nrow(df), 2)`%) have no female speakers or chairs.


### All-female events

```{r}
all_female = filter(df, male.speakers == 0 & male.chair == 0)
```


`r nrow(filter(df, male.speakers == 0))` (`r 100*round(nrow(filter(df, male.speakers == 0))/nrow(df),2)`%) have no male speakers and `r nrow(all_female)` (`r 100*round(nrow(filter(df, male.speakers == 0 & male.chair == 0))/nrow(df), 2)`%).

What are the all-female events about?

```{r}
all_female$title
colSums(all_female[,7:12])>0
```
